Numerical simulation of bubble deformation \\ and breakup under simple linear shear flows

### 2022-12-27 Bubble-deformation.tex
figures/Fig11-Critical-Re1_cubic.jpg

Figure~\ref{fig:SchemAndGrid}(a) shows a schematic of the computational system
for our studies of a bubble (or drop) in shear flow.  The computational domain
consists of a three-dimensional rectangular domain of $L(x)$ (length) $\times$
$W(y)$ (width) $\times$ $H(z)$ (height).  The size of $L(x)$, $W(y)$ and $H(z)$
was determined after consideration of numerical result sensitivity to domain
size; numerical studies of domain-size dependence are presented in
Section~\ref{sec:DomGrdSize}.  All computational results that follow were
obtained from numerical solutions of the three-dimensional governing equations
for gas-liquid/liquid-liquid flows.  Computations are initialized with a
spherical bubble set at the center of the domain that is then subjected to a
linear shear flow generated by the motion of the top and bottom plates, which 
have constant velocity $+V$ and $-V$, respectively.  In the interior of the
domain, the initial velocity condition is assumed to be a simple linear profile
and periodic boundary conditions are imposed along the $x$ and $y$ directions.
Mathematically, the initial and boundary conditions are described as
follows:
\begin{eqnarray}
	\phi(x,y,z,0)=\sqrt{(x-\frac{L}{2})^{2}+(y-\frac{W}{2})^{2}+
	(z-\frac{H}{2})^{2}}-R \label{IC_BC} \\
	\bmu(x,y,z,0)=\left( \begin{array}{c}
		\frac{2V}{H}(z-\frac{H}{2}) \\ 0 \\ 0 
	\end{array} \right) \nonumber \\
	\phi(x+L,y,z,t)=\phi(x,y,z,t) \hspace{20pt}
	\phi(x,y+W,z,t)=\phi(x,y,z,t) \nonumber \\
	\bmu(x,y,H,t)=\left( \begin{array}{c}
		                V \\ 0 \\ 0 
	\end{array} \right)  \hspace{20pt}
	\bmu(x,y,0,t)=\left( \begin{array}{c}
		                -V \\ 0 \\ 0 
	\end{array} \right)  \nonumber \\
	\bmu(x+L,y,z,t)=\bmu(x,y,z,t) \hspace{20pt} 
	\bmu(x,y+W,z,t)=\bmu(x,y,z,t). \nonumber
\end{eqnarray}
$\phi(x,y,z,t)$ is a material indicator function which is positive 
in the ``matrix'' fluid and negative in the bubble (or drop) fluid.
$\bmu(x,y,z,t)$ is the velocity.

@@

Table~\ref{tab:CaRecComparison} 
lists, for representative $Ca$ values,
the corresponding critical Reynolds number, $Re_{c}$, for shear-induced 
bubble breakup for a variety of flow
conditions.  The results indicate that sufficiently large shear forces are
required for bubble breakup even for large Capillary numbers. The combined
effects of very small density and viscosity ratios 
($\lambda$ = $\eta$ = 1.0 $\times$ $10^{-3}$) 
act synergistically in deformation and breakup processes.
We conducted a linear least squares best fit analysis (linear vs. 
power law vs. exponential models) of the data that we 
list in Table~\ref{tab:CaRecComparison} and have found that the following
exponential model is the best fit:
%  Re = \frac{\rho_{b}UR}{\mu_{b}}, \quad
%  Ca = \frac{\mu_{b}U}{\sigma}, \quad
\begin{eqnarray}
Re_{c}=138.3 e^{-1.41 (Ca)} \label{bestfit}
\end{eqnarray}
(\ref{bestfit}) was derived from the linear least squares best fit
of the log of $Re_{c}$:
\begin{eqnarray}
	\log(Re_{c})=4.9-1.41 (Ca)  \label{bestfitraw}
\end{eqnarray}
%% variance for X^T X c = b
%% (S/(n-m)) (X^T X)^{-1}
%% standard deviation: sqrt(4.2E-3)=.07 and sqrt(8.4E-3)=0.09
The predicted variances for the parameters 
``$4.9$'' and ``$-1.41$'' which appear in (\ref{bestfitraw}) are
$4.2E-3$ and $8.4E-3$ respectively.  In other words, even with just
four data points, we have computed a reasonable best fit curve.

In Figure~\ref{fig:CaRecFit} we plot our best fit curve and make the
hypothesis that a given $(Ca,Re)$ pair will result in bubble break up
if the point $(Ca,Re)$ is above the critical curve, 
$0.3<Ca<1$, and the initial/boundary conditions are given
by (\ref{IC_BC}).

@@

%% variance for X^T X c = b
%% (S/(n-m)) (X^T X)^{-1}
\begin{figure}%[h!]
  \centering
  \includegraphics[width=\textwidth]{Fig12-Critical-Re}
  \caption{A plot of the critical Reynolds number versus Capillary number
	for the data listed in Table \ref{tab:CaRecComparison}.  The
	linear least squares best fit for the $\log(Re_{c})$ is
	$\log(Re_{c})=4.9-1.41(Ca)$ (i.e. $Re_{c}=138.3 e^{-1.41 Ca}$)
	with a predicted variance
	for the parameters ``$4.9$'' and ``$-1.41$'' being
	$4.2E-3$ and $8.4E-3$ respectively.  Even for the
	given data that we computed, it is suggested that 
	a ``bubble break-up map''
	can be derived, in which for the conditions that
	$\lambda$ = $\eta$ = 1.0 $\times$ $10^{-3}$, $0.3<Ca<1$, and
        the initial/boundary conditions are given by (\ref{IC_BC}),	
	one
	can predict whether a given $(Ca,Re)$ pair will result in shear
	induced bubble break-up or not.  
	   }
  \label{fig:CaRecFit}
\end{figure}
%

>>>>

Figure~\ref{fig:SchemAndGrid}(a) shows a schematic of the computational system
for our studies of a bubble (or drop) in shear flow.  The computational domain
consists of a three-dimensional rectangular domain of $L(x)$ (length) $\times$
$W(y)$ (width) $\times$ $H(z)$ (height).  The size of $L(x)$, $W(y)$ and $H(z)$
was determined after consideration of numerical result sensitivity to domain
size; numerical studies of domain-size dependence are presented in
Section~\ref{sec:DomGrdSize}.  All computational results that follow were
obtained from numerical solutions of the three-dimensional governing equations
for gas-liquid/liquid-liquid flows.  Computations are initialized with a
spherical bubble set at the center of the domain that is then subjected to a
linear shear flow generated by the motion of the top and bottom plates, which 
have constant velocity $+V$ and $-V$, respectively.  In the interior of the
domain, the initial velocity condition is assumed to be a simple linear profile
and periodic boundary conditions are imposed along the $x$ and $y$ directions.
Mathematically, the initial and boundary conditions are described as
follows:
\begin{eqnarray}
	\phi(x,y,z,0)=\sqrt{(x-\frac{L}{2})^{2}+(y-\frac{W}{2})^{2}+
	(z-\frac{H}{2})^{2}}-R \label{IC_BC} \\
	\bmu(x,y,z,0)=\left( \begin{array}{c}
		\frac{2V}{H}(z-\frac{H}{2}) \\ 0 \\ 0 
	\end{array} \right) \nonumber \\
	\phi(x+L,y,z,t)=\phi(x,y,z,t) \hspace{20pt}
	\phi(x,y+W,z,t)=\phi(x,y,z,t) \nonumber \\
	\bmu(x,y,H,t)=\left( \begin{array}{c}
		                V \\ 0 \\ 0 
	\end{array} \right)  \hspace{20pt}
	\bmu(x,y,0,t)=\left( \begin{array}{c}
		                -V \\ 0 \\ 0 
	\end{array} \right)  \nonumber \\
	\bmu(x+L,y,z,t)=\bmu(x,y,z,t) \hspace{20pt} 
	\bmu(x,y+W,z,t)=\bmu(x,y,z,t) \nonumber
\end{eqnarray}
$\phi(x,y,z,t)$ is a material indicator function which is positive 
in the ``matrix'' fluid and negative in the bubble (or drop) fluid.
$\bmu(x,y,z,t)$ is the velocity.

@@

Table~\ref{tab:CaRecComparison} 
lists, for representative $Ca$ values,
the corresponding critical Reynolds number, $Re_{c}$, for shear-induced 
bubble breakup.  The data in 
Table~\ref{tab:CaRecComparison} 
corresponds to $\lambda$ = $\eta$ = 1.0 $\times$ $10^{-3}$, $0.3<Ca<1$, and
the initial/boundary conditions are given by (\ref{IC_BC}).
The results in 
Table~\ref{tab:CaRecComparison}
indicate that sufficiently large shear forces are
required for bubble breakup even for large Capillary numbers. 
In Figure~\ref{fig:CaRecFit} we plot the cubic interpolant of the
data given in 
Table~\ref{tab:CaRecComparison}
and make the
hypothesis that given a new data point, $(Ca,Re)$, 
shear induced bubble break up
will occur if the point $(Ca,Re)$ is above the given critical curve, 
and the bubble will not break if the $(Ca,Re)$ pair is below the 
critical curve.

@@

%% variance for X^T X c = b
%% (S/(n-m)) (X^T X)^{-1}
\begin{figure}%[h!]
  \centering
  \includegraphics[width=\textwidth]{Fig11-Critical-Re1_cubic}
  \caption{A plot of the critical Reynolds number versus Capillary number
	for the data listed in Table~\ref{tab:CaRecComparison}.  
	$\lambda$ = $\eta$ = 1.0 $\times$ $10^{-3}$, $0.3<Ca<1$, and
        the initial/boundary conditions are given by (\ref{IC_BC}).
	Given a new data point, $(Ca,Re)$, one
	can reliably predict whether the 
	given $(Ca,Re)$ pair will result in shear
	induced bubble break-up or not.  
	   }
  \label{fig:CaRecFit}
\end{figure}
%
